Linear Vlasov Theory in the Shearing Sheet Approximation \\ with Application to the Magneto-Rotational Instability

### 2015-01-02 paper.tex
Rephrased a few sentences
... in an effort to improve clarity.

The equation of motion \labelcref{eq:particle-eom} is generated by the
Hamiltonian\footnote{Here and in the following we will refer to both
  $\mathcal{H}$ and $2\mathcal{H}$ as the Hamiltonian. The same goes for any
  energy-like quantity.}
\begin{equation}
  \label{eq:hamiltonian}
  2\mathcal{H} = p_x^2 + {(p_y - S_z x)}^2 + {(p_z + S_y x)}^2
  - q\Omega S_z x^2 + \vfreq^2 z^2,
\end{equation}
see \cref{app:unperturbed-hamiltonian} for a derivation. Expressing this in
terms of the particle velocity $\vec{v}=\partial\mathcal{H}/\partial\vec{p}$
yields
\begin{equation}
  \label{eq:energy-integral}
  2\mathcal{E} = \vec{v}^2 - q\Omega S_z x^2 + \vfreq^2 z^2,
\end{equation}
which we will refer to as the energy integral or simply the energy. According
to Jeans' theorem, any distribution function that depends only on the
integrals of motion is in equilibrium. Thus, any $f(\mathcal{E})$ is in
equilibrium. However, any such distribution function describes an equilibrium
in which the plasma is not rotationally supported but pressure-supported
against the tidal force: there is no mean flow and the density is non-uniform.
An equilibrium distribution function that is compatible with the underlying
assumptions of the shearing sheet is obtained as follows.

@@

In this section we study modes whose wave vector is aligned with the magnetic
field, which in turn is aligned with the rotation axis. The ion response
tensor in this case is given by
\begin{equation}
  \label{eq:parallel-response}
  \mathbf{\Lambda} =
  \frac{1}{2}\sum_\pm\Lambda^{}_\pm
  (\vec{e}_1\pm i\vec{e}_2)(\vec{e}_1\mp i\vec{e}_2)
  - \frac{\omega^2}{k_z^2\vt^2} W\left(\frac{\omega}{|k_z|\vt}\right)
  \vec{e}_3\vec{e}_3,
\end{equation}
with perpendicular components
\begin{equation}
  \label{eq:circular-components}
  \Lambda_\pm = \frac{\omega}{\omega\pm\omg}
  \left[1 - W\left(\frac{\omega\pm\omg}{|k_z|\vt}\right)\right].
\end{equation}
We note that $\ex\ne\vec{e}_1$ and $\ey\ne\vec{e}_2$ in general even for
parallel modes. This is because the orthonormal basis
$(\vec{e}_1,\vec{e}_2,\vec{e}_3)$ with $\vec{e}_3=\vomg/\omg$ is taken to be
right-handed (see \cref{sec:plasma-response}), and whether or not the
gyro-frequency vector $\vomg$ defined in \cref{eq:gyro-frequency-vector} is
aligned with the angular velocity vector $\vec{\Omega}=\Omega\ez$ depends on
the sign of $S_z=\omc{}b_z+2\Omega$. In this context we also note that the
gyration frequency $\omg>0$ throughout this text.\footnote{Unless of course
  $\omg^2<0$, in which case much of our analysis is invalidated, see the
  discussion in \cref{sec:delta}.} The cyclotron frequency $\omc=eB/m$ can
have either sign, but this depends only on the sign of the charge, \emph{not}
on the direction of the magnetic field.

@@

In \cref{sec:guiding-center} we have shown that we recover the kinetic MHD
result obtained by \citet{Quataert2002} in the guiding center limit
$\omc\to\infty$. For parallel modes it is easy to calculate the first order
correction for finite cyclotron frequency.\footnote{Note that for parallel
  modes the MHD and kinetic MHD dispersion relations are identical} In order
to do so, let us assume the ordering
\begin{equation}
  \label{eq:ordering}
  \omega\sim k_z\va\sim\Omega\sim\epsilon\beta\Omega
  \sim\epsilon^2\omc,
\end{equation}
where the ordering parameter $\epsilon\ll1$. We thus assume that the plasma is
hot but not too hot in the sense that $1\ll\beta\ll\omc/\Omega$. The
dispersion tensor to first order in $\epsilon$ is then given by
\begin{equation}
  \label{eq:gyro-viscous-drel}
  \mathbf{D}_\perp = (k_z^2\va^2 - \omega^2)\mathbf{1} - i\omega\left(
    2\Omega - \frac{k_z^2\vt^2}{\omc b_z}
  \right)\ez\times\mathbf{1} - 2q\Omega^2\ey\ey.
\end{equation}
Setting its determinant equal to zero yields the dispersion relation
\begin{equation}
  \omega^4
  - \omega^2[\kappa^2 + 2k_z^2\va^2\left(1 - \beta\Omega/\omc b_z\right)]
  + k_z^2\va^2(k_z^2\va^2 - 2q\Omega^2) = 0.
\end{equation}
The lowest order correction to the standard MHD dispersion relation is thus of
order $\beta\Omega/\omc\ll{}1$. This is the magnitude of the gyro-viscous
force relative to the magnetic tension force in the Braginskii equations, a
fluid closure for weakly collisional plasmas that includes FLR
effects.\footnote{In this context we remark that the gyro-viscous force in the
  Braginskii equations does in fact \emph{not} depend on the collision
  frequency.} \citet{Ferraro2007} provides a numerical analysis of the MRI
with the gyro-viscous force taken into account. Our results confirm that
gyro-viscosity is indeed the dominant FLR effect in a hot plasma, not the Hall
effect.

@@

We have for the first time derived the general linear conductivity tensor for
a collisionless plasma in the shearing sheet approximation. This accounts for
the complete linear dynamics of kinetic ions \emph{and} kinetic electrons with
finite Larmor radius/cyclotron frequency. Our calculation has leveraged the
extensive literature on analogous calculations for uniform plasmas
\citep{Ichimaru1973}. In particular, we have shown that by a suitable choice
of velocity space coordinates, the linear conductivity tensor for a
collisionless plasma in the shearing sheet approximation can be expressed
using standard results for a uniform plasma
(\cref{sec:conductivity-tensor,app:conductivity-tensor}). In doing so, our
primary assumptions are charge-neutrality, axisymmetric perturbations, and no
vertical stratification. The charge-neutral approximation is made both for its
applicability to realistic accretion disk dynamics and because the shearing
sheet approximation is locally Galilean invariant, not Lorentz invariant.
Galilean invariance implies charge-neutrality once the displacement current is
dropped \citep{Grad1966}.

@@

The situation is more complicated if there is a toroidal magnetic field. In
this case there is a region of parameter space where circular orbits are
linearly stable even though they do not minimize the energy. To see this, let
us assume (as in the bulk of the main text) that there is no stratification
($\Psi=\mathrm{const}$). In this case, the solubility condition
\cref{eq:solubility-condition} reduces to
\begin{equation}
  \label{eq:solubility-no-stratification}
  \omega^2(\omega^2 - \omg^2) = 0,
\end{equation}
where the square of the gyration frequency $\omg$ is given by
\begin{equation}
  \label{eq:global-gyration-freq}
  \omg^2 = S_z\left(S_z + r\pder{\Omega}{r}\right) + S_\varphi^2.
\end{equation}
This is the global analogue of the local gyration frequency defined in
\cref{eq:gyro-frequency} in the main text.
\Cref{eq:solubility-no-stratification} with \cref{eq:global-gyration-freq}
implies that circular orbits are linearly stable provided that
\begin{equation}
  \Delta < 1 + S_\varphi^2/S_z^2,
\end{equation}
which does \emph{not} coincide with the nonlinear stability boundary
$\Delta<1$. In this context we remark that according to a theorem due to
\citet{Bloch1994}, a relative equilibrium\footnote{Circular orbits are
  relative equilibria.} that is \emph{not} nonlinearly stable is destabilized
when a small amount of dissipation is added to the system. We are thus led to
conclude that circular orbits in realistic systems are always unstable if
$\Delta>1$, regardless of whether or not there is a toroidal magnetic field.

@@

The electron velocity $\ue$ plays a more fundamental role in Vlasov-fluid
theory than may be apparent at first sight. Differentiating
\cref{eq:charge-neutrality} with respect to time yields after a little bit of
algebra the charge continuity equation
\begin{equation}
  \label{eq:charge-continuity}
  \pder{n_e}{t} + \nabla\cdot(n_e\vec{u}_e) = 0.
\end{equation}
With the barotropic assumption, eliminating the electric field between Ohm's
law and Faraday's law yields the induction equation
\begin{equation}
  \label{eq:induction}
  \pder{\vec{B}}{t} = \nabla\times(\vec{u}_e\times\vec{B}).
\end{equation}
The magnetic field is thus frozen into the electron fluid. The form of
\cref{eq:charge-continuity,eq:induction} suggests that in linear theory we
introduce the Lagrangian displacement through
\begin{equation}
  \label{eq:delta-ue}
  \delta\ue + \vec{\xi}\cdot\nabla\ue =
  \left(\pder{}{t} + \ue\cdot\nabla\right)\vec{\xi}
\end{equation}
cf.\ \citet{Lynden-Bell1967,Friedman1978}. The linearized
\cref{eq:charge-continuity,eq:induction} are then easily integrated to yield
\begin{equation}
  \label{eq:delta-ne}
  \delta n_e + \nabla\cdot(n_e\vec{\xi}) = 0
\end{equation}
and
\begin{equation}
  \label{eq:delta-B}
  \delta\vec{B} = \nabla\times(\vec{\xi}\times\vec{B}).
\end{equation}
We stress that \cref{eq:delta-ue,eq:delta-ne,eq:delta-B} hold for general
equilibria.

>>>>

The equation of motion \labelcref{eq:particle-eom} is generated by the
Hamiltonian\footnote{Here and in the following we will refer to both
  $\mathcal{H}$ and $2\mathcal{H}$ as the Hamiltonian. The same goes for any
  energy-like quantity.}
\begin{equation}
  \label{eq:hamiltonian}
  2\mathcal{H} = p_x^2 + {(p_y - S_z x)}^2 + {(p_z + S_y x)}^2
  - q\Omega S_z x^2 + \vfreq^2 z^2,
\end{equation}
see \cref{app:unperturbed-hamiltonian} for a derivation. Expressing this in
terms of the particle velocity $\vec{v}=\partial\mathcal{H}/\partial\vec{p}$
yields
\begin{equation}
  \label{eq:energy-integral}
  2\mathcal{E} = \vec{v}^2 - q\Omega S_z x^2 + \vfreq^2 z^2,
\end{equation}
which we will refer to as the energy integral or simply the energy. According
to Jeans' theorem, any distribution function that depends only on the
integrals of motion is in equilibrium. Thus, any $f(\mathcal{E})$ is in
equilibrium. However, any such distribution function describes an equilibrium
in which the plasma is not rotationally supported but pressure-supported
against the tidal force: there is no mean flow and the density is non-uniform
in $x$. An equilibrium distribution function that is compatible with the
underlying assumptions of the shearing sheet is obtained as follows.

@@

In this section we study modes whose wave vector is aligned with the magnetic
field, which in turn is aligned with the rotation axis. The ion response
tensor in this case is given by
\begin{equation}
  \label{eq:parallel-response}
  \mathbf{\Lambda} =
  \frac{1}{2}\sum_\pm\Lambda^{}_\pm
  (\vec{e}_1\pm i\vec{e}_2)(\vec{e}_1\mp i\vec{e}_2)
  - \frac{\omega^2}{k_z^2\vt^2} W\left(\frac{\omega}{|k_z|\vt}\right)
  \vec{e}_3\vec{e}_3,
\end{equation}
with perpendicular components
\begin{equation}
  \label{eq:circular-components}
  \Lambda_\pm = \frac{\omega}{\omega\pm\omg}
  \left[1 - W\left(\frac{\omega\pm\omg}{|k_z|\vt}\right)\right].
\end{equation}
At this point we note that even in the case of a purely vertical magnetic
field ($\vec{B}=B_z\ez$), the coordinate basis
$(\vec{e}_1,\vec{e}_2,\vec{e}_3)$ as defined in \cref{sec:plasma-response}
does not in general coincide with $(\ex,\ey,\ez)$. This is because
$\vec{e}_3=\vomg/\omg$, and whether the gyro-frequency vector $\vomg$ is
aligned or anti-aligned with the angular frequency vector
$\vec{\Omega}=\Omega\ez$ depends on the sign of $S_z=\omc{}b_z+2\Omega$. In
this context we also note that the gyration frequency $\omg>0$ throughout this
text.\footnote{Unless of course $\omg^2<0$, in which case much of our analysis
  is invalidated, see the discussion in \cref{sec:delta}.} The cyclotron
frequency $\omc=eB/m$ can have either sign, but this depends only on the sign
of the charge, \emph{not} on the direction of the magnetic field.

@@

In \cref{sec:guiding-center} we have shown that we recover the kinetic MHD
result obtained by \citet{Quataert2002} in the guiding center limit
$\omc\to\infty$. For parallel modes it is easy to calculate the first order
correction for finite cyclotron frequency.\footnote{Note that for parallel
  modes the MHD and kinetic MHD dispersion relations are identical} In order
to do so, let us assume the ordering
\begin{equation}
  \label{eq:ordering}
  \omega\sim k_z\va\sim\Omega\sim\epsilon\beta\Omega
  \sim\epsilon^2\omc,
\end{equation}
where the ordering parameter $\epsilon\ll1$. We thus assume that the plasma is
hot but not too hot in the sense that $1\ll\beta\ll\omc/\Omega$. The
perpendicular dispersion tensor to first order in $\epsilon$ is then given by
\begin{equation}
  \label{eq:gyro-viscous-drel}
  \mathbf{D}_\perp = (k_z^2\va^2 - \omega^2)(\mathbf{1} - \ez\ez)
  - i\omega\left(
    2\Omega - \frac{k_z^2\vt^2}{\omc b_z}
  \right)\ez\times\mathbf{1} - 2q\Omega^2\ey\ey.
\end{equation}
Setting its determinant equal to zero yields the dispersion relation
\begin{equation}
  \omega^4
  - \omega^2[\kappa^2 + 2k_z^2\va^2\left(1 - \beta\Omega/\omc b_z\right)]
  + k_z^2\va^2(k_z^2\va^2 - 2q\Omega^2) = 0.
\end{equation}
The lowest order correction to the standard MHD dispersion relation is thus of
order $\beta\Omega/\omc\ll{}1$. This is the magnitude of the gyro-viscous
force relative to the magnetic tension force in the Braginskii equations, a
fluid closure for weakly collisional plasmas that includes FLR
effects.\footnote{In this context we remark that the gyro-viscous force in the
  Braginskii equations does in fact \emph{not} depend on the collision
  frequency.} \citet{Ferraro2007} provides a numerical analysis of the MRI
with the gyro-viscous force taken into account. Our results confirm that
gyro-viscosity is indeed the dominant FLR effect in a hot plasma, not the Hall
effect.

@@

We have for the first time derived the general linear conductivity tensor for
a collisionless plasma in the shearing sheet approximation. This accounts for
the complete linear dynamics of kinetic ions \emph{and} kinetic electrons with
finite Larmor radius/cyclotron frequency. Our calculation has leveraged the
extensive literature on analogous calculations for uniform plasmas
\citep{Ichimaru1973}. In particular, we have shown that by a suitable choice
of velocity space coordinates, the linear conductivity tensor for a
collisionless plasma in the shearing sheet approximation can be expressed
using standard results for a uniform plasma
(\cref{sec:conductivity-tensor,app:conductivity-tensor}). In doing so, our
primary assumptions are charge-neutrality, axisymmetric perturbations, and no
vertical stratification. The charge-neutral approximation is made both for its
applicability to realistic accretion disk dynamics and because the shearing
sheet approximation in its standard form relies on Galilean invariance, which
in turn necessitates charge-neutrality once the displacement current is
dropped \citep[see e.g.][]{Grad1966}.

@@

The situation is more complicated if there is a toroidal magnetic field. In
this case there is a region of parameter space where circular orbits are
linearly stable even though they do not minimize the energy. To see this, let
us assume (as in the bulk of the main text) that there is no stratification
($\Psi=\mathrm{const}$). In this case, the solubility condition
\cref{eq:solubility-condition} reduces to
\begin{equation}
  \label{eq:solubility-no-stratification}
  \omega^2(\omega^2 - \omg^2) = 0,
\end{equation}
where the square of the gyration frequency $\omg$ is given by
\begin{equation}
  \label{eq:global-gyration-freq}
  \omg^2 = S_z\left(S_z + r\pder{\Omega}{r}\right) + S_\varphi^2.
\end{equation}
This is the global analogue of the local gyration frequency defined in
\cref{eq:gyro-frequency} in the main text.
\Cref{eq:solubility-no-stratification} with \cref{eq:global-gyration-freq}
implies that circular orbits are linearly stable provided that
\begin{equation}
  \Delta < 1 + S_\varphi^2/S_z^2,
\end{equation}
which does \emph{not} coincide with the nonlinear stability boundary
$\Delta<1$. In this context we remark that according to a theorem due to
\citet{Bloch1994}, a relative equilibrium\footnote{Circular orbits are
  relative equilibria.} that is linearly but \emph{not} nonlinearly stable
becomes linearly unstable when a small amount of dissipation is added to the
system. We are thus led to conclude that circular orbits in realistic systems
are always unstable if $\Delta>1$, regardless of whether or not there is a
toroidal magnetic field.

@@

The electron velocity $\ue$ plays a more fundamental role in Vlasov-fluid
theory than may be apparent at first sight. Differentiating
\cref{eq:charge-neutrality} with respect to time yields after a little bit of
algebra the continuity equation
\begin{equation}
  \label{eq:charge-continuity}
  \pder{n_e}{t} + \nabla\cdot(n_e\vec{u}_e) = 0.
\end{equation}
With the barotropic assumption, eliminating the electric field between Ohm's
law and Faraday's law yields the induction equation
\begin{equation}
  \label{eq:induction}
  \pder{\vec{B}}{t} = \nabla\times(\vec{u}_e\times\vec{B}).
\end{equation}
The magnetic field is thus frozen into the electron fluid. The form of
\cref{eq:charge-continuity,eq:induction} suggests that in linear theory we
introduce the Lagrangian displacement through
\begin{equation}
  \label{eq:delta-ue}
  \delta\ue + \vec{\xi}\cdot\nabla\ue =
  \left(\pder{}{t} + \ue\cdot\nabla\right)\vec{\xi}
\end{equation}
cf.\ \citet{Lynden-Bell1967,Friedman1978}. The linearized
\cref{eq:charge-continuity,eq:induction} are then easily integrated to yield
\begin{equation}
  \label{eq:delta-ne}
  \delta n_e + \nabla\cdot(n_e\vec{\xi}) = 0
\end{equation}
and
\begin{equation}
  \label{eq:delta-B}
  \delta\vec{B} = \nabla\times(\vec{\xi}\times\vec{B}).
\end{equation}
We stress that \cref{eq:delta-ue,eq:delta-ne,eq:delta-B} hold for general
equilibria.
